---
title: "Out-of-Core Spectronaut Workflow with matter and Arrow"
author: "Tony Wu (<wu.anthon@northeastern.edu>)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Out-of-Core Spectronaut Workflow with matter and Arrow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

This vignette demonstrates the matter-backed workflow in `MSstatsBig`:

1. Convert a Spectronaut export to a **matter bundle** with
   `bigSpectronauttoMSstatsFormat(convert_matter = TRUE)`.
2. Summarize per protein in parallel with `dataProcessBig()`.

Compared to the standard `MSstatsBig` → `MSstats::dataProcess` flow, this path
keeps all per-protein data on disk (in a `matter_list`) and only loads one
protein at a time during summarization. Peak RAM is bounded by the largest
single protein, not by the total dataset size.

## When to use this workflow

The standard workflow is fine when the *post-preprocessing* output fits
comfortably in RAM. Use the matter-backed workflow when:

- The dataset has tens of thousands of proteins and hundreds of runs (DIA
  scale).
- You want to run summarization repeatedly (e.g., re-running with different
  `n_top_feature` or imputation settings) without re-converting.
- You need parallel summarization where workers should not each receive a
  copy of the full feature-level table.

## Pipeline

```
spectronaut_export.csv
        │
        │  bigSpectronauttoMSstatsFormat(convert_matter = TRUE)
        │    ├── reduce + clean (existing path)
        │    ├── arrow preprocessing (top-N feature selection)
        │    └── .convertArrowToMatterList:
        │         ├── lazy log2 transform
        │         ├── per-run median (Arrow streaming aggregate)
        │         ├── repartition by ProteinName
        │         └── per-protein matter_list write (incremental)
        ▼
   matter bundle/
        ├── intensities.bin       ← matter_list, one slot per protein
        ├── run_stats.parquet     ← per-run median + n_obs
        ├── proteins.parquet      ← k → ProteinName, n_features, n_runs
        ├── features.parquet      ← per-protein feature ordering + metadata
        ├── runs.parquet          ← row ordering + Condition / BioReplicate
        └── manifest.rds          ← canonical run order, log2 flag
        │
        │  dataProcessBig()
        │    └── per protein in parallel:
        │         ├── load intensities[[k]] from disk
        │         ├── equalize-medians normalize (vectorized)
        │         ├── top-N feature selection (post-norm)
        │         ├── pack to V3 format
        │         └── MSstatsSummarizeSingleTMPV2 (impute + TMP)
        ▼
   ProteinLevelData.rds            ← single MSstats-format file
```

## Setup

```{r load_package}
library(MSstatsBig)
library(arrow)
library(matter)
```

## 1. Convert Spectronaut export to a matter bundle

```{r convert}
work_dir <- tempdir()
setwd(work_dir)

result <- bigSpectronauttoMSstatsFormat(
    input_file        = system.file("extdata", "spectronaut_input.csv",
                                    package = "MSstatsBig"),
    output_file_name  = "preprocessed.csv",
    backend           = "arrow",
    intensity         = "F.NormalizedPeakArea",
    filter_by_qvalue  = TRUE,
    qvalue_cutoff     = 0.01,
    max_feature_count = 20,             # top-N at conversion time
    convert_matter    = TRUE,           # write the matter bundle
    matter_output_dir = "spectronaut_bundle"
)
```

`result` is a list with two elements when `convert_matter = TRUE`:

- `result$data` — the Arrow lazy reference to the preprocessed CSV (same as
  the non-matter path).
- `result$matter_bundle` — file paths inside the bundle directory.

```{r inspect_bundle}
result$matter_bundle
```

You can inspect the bundle pieces directly:

```{r inspect_metadata}
proteins_meta <- read_parquet(result$matter_bundle$proteins)
runs_meta     <- read_parquet(result$matter_bundle$runs)
run_stats     <- read_parquet(result$matter_bundle$run_stats)
manifest      <- readRDS(result$matter_bundle$manifest)

head(proteins_meta)
head(run_stats)
manifest$canonical_run_order
manifest$log2_intensity     # TRUE — log2 was applied during conversion
```

The `intensities.bin` backing file holds, for each protein *k*, a
`(n_runs × n_features)` matrix flattened column-major as `double`:

```{r inspect_intensities}
intensities <- matter_list(
    NULL,
    type     = "double",
    path     = result$matter_bundle$intensities,
    lengths  = as.integer(proteins_meta$n_features) * manifest$n_runs,
    names    = proteins_meta$ProteinName,
    readonly = TRUE)

# Pull one protein's matrix from disk on demand.
k        <- 1L
n_runs   <- manifest$n_runs
n_feat_k <- proteins_meta$n_features[k]
mat_k    <- matrix(intensities[[k]], nrow = n_runs, ncol = n_feat_k)
dim(mat_k)
mat_k[1:5, 1:min(5, ncol(mat_k))]
```

## 2. Summarize with `dataProcessBig`

```{r data_process}
plv <- dataProcessBig(
    matter_dir      = result$matter_bundle$output_dir,
    output_file     = "ProteinLevelData.rds",
    n_top_feature   = 20L,           # post-normalization top-N
    impute          = TRUE,          # AFT survival imputation
    censored_symbol = "NA",          # NAs treated as below detection
    remove50missing = FALSE,
    numberOfCores   = 4L,
    verbose         = TRUE
)

head(plv)
```

`plv` is a `data.table` matching the MSstats `ProteinLevelData` schema:

```
Protein, LABEL, RUN, originalRUN, LogIntensities,
GROUP, SUBJECT, TotalGroupMeasurements,
NumMeasuredFeature, MissingPercentage, more50missing, NumImputedFeature
```

It is also written to the path you passed via `output_file` (RDS by default;
parquet or CSV when the extension implies it).

## 3. Hand off to MSstats `groupComparison`

`dataProcessBig` produces only the `ProteinLevelData` slot. `MSstats::groupComparison`
expects a list with `ProteinLevelData` (and optionally `FeatureLevelData`). You
can wrap it for compatibility:

```{r group_comparison}
library(MSstats)

summarized <- list(
    ProteinLevelData = as.data.frame(plv),
    FeatureLevelData = NULL,             # not produced by dataProcessBig
    SummaryMethod    = "TMP"
)

# Example contrast for the Spectronaut sample (adjust to your conditions)
groups <- unique(plv$GROUP)
comparison <- matrix(c(-1, 1), nrow = 1, byrow = TRUE,
                     dimnames = list("contrast1", groups[1:2]))

results <- groupComparison(
    contrast.matrix = comparison,
    data            = summarized,
    use_log_file    = FALSE)

head(results$ComparisonResult)
```

## Memory profile at a glance

| Step                                  | Peak RAM held by main process     |
|---------------------------------------|------------------------------------|
| Per-run median scan (Arrow)           | one row per run                    |
| Distinct features scan (Arrow)        | one row per (protein, feature)     |
| `arrow::write_dataset` repartition    | one Arrow batch (streaming)        |
| `matter_list` pre-allocation          | empty backing file, no data        |
| Per-protein write loop                | one protein's matrix at a time     |
| `dataProcessBig` summarization        | one protein per worker             |
| ProteinLevelData assembly             | the final result table             |

The full preprocessed table is **never** materialized as a data.frame in
memory.

## Re-running summarization on the same bundle

A useful property of the bundle: it's pure data, no MSstats state. You can
re-run summarization with different parameters without re-converting:

```{r rerun}
# Try a different feature count and imputation off
plv_no_impute <- dataProcessBig(
    matter_dir    = result$matter_bundle$output_dir,
    output_file   = "ProteinLevelData_noimpute.rds",
    n_top_feature = 10L,
    impute        = FALSE,
    numberOfCores = 4L)
```

## Assumptions and limitations

The current matter workflow assumes:

- **No duplicate (Run, Feature) pairs** — one intensity per cell.
- **No fractions** — `runs.parquet` has one row per Run.
- **Unlabeled experiment** — single `IsotopeLabelType`. SILAC/iTRAQ datasets
  will load but the per-protein matrix layout treats each label as an
  independent feature column rather than pairing them.
- **`matter` and `arrow` packages installed**.

`run_stats.parquet` was computed across all observations (not filtered by
reference label), so the equalize-medians normalization is exact for
unlabeled experiments and approximate for labeled ones until the SILAC path
lands.

## Session info

```{r session_info, eval = TRUE}
sessionInfo()
```
